---
title: "Regression Trees"
format:
    html:
        theme: darkly
        max-width: 800px
        fontsize: 12pt
subtitle: "Lesson 1.5, Applied Machine Learning"
---

[Course Homepage](https://politicalscience.github.io/ml)

## Table of Contents {#contents}

1.  [Regression Trees](#trees)
2.  [Regression Trees in R](#r)

------------------------------------------------------------------------

Remember to load tidyverse. We will also need the package tree.

```{r, message = FALSE}
library(tidyverse)
library(tree)
```

Let us also load the dataset we will be using for these examples (feel free to load your own dataset)

```{r, message = FALSE}
df <- read_csv("voctaxdata.csv")
```

------------------------------------------------------------------------

[Table of Contents](#contents) \| [Course Homepage](https://politicalscience.github.io/ml)

# Regression Trees {#trees}

Regression Trees are an alternative way to obtain estimations of $y$ from $x$. Instead of modelling how $y$ changes with every unit change in $x$, regression trees instead use stratification:

1.  Tree-based methods will divide the independent variable $x$ into 2 regions, splitting $x$ at some value $s$
    -   For example, if we had an $x$ variable with values between $0$ and $100$, we could split the variable at $x=50$ to create two regions.
2.  Calculate the mean outcome value $y$ in each region of $x$
    -   For example, if we had split $x$ at $x=50$, we would have 2 regions of $x$: 0 to 49, and 50 to 100. We would calculate the average $y$ value of each region.
3.  Predict $y$, based on the mean $y$ of the region which the observation falls into.
    -   The region in which the observation falls into depends on its value of $x$. For example, if we follow the splits above, if an observation had $x=5$, it would be classified in the region of 0 to 49, and its prediction would be the mean $y$ value of that region's observations.

<br />

If we have to split $x$ into regions, at what threshold of $x$ should the regions be split at?

-   For example, if we had an $x$ variable with values between $0$ and $100$, we could split the variable at $x=50$, or $x=20$, or really any value of $x$.

To determine at which threshold to split $x$, the computer will find the optimal $x$ threshold for splitting based on which value reduces the residual sum of squares (RSS) the most.

$$
RSS = \sum\limits_{i=1}^{n} (y_i - \hat{y}_i)
$$

-   Essentially, the computer tests all possible $x$ values in which we could split by, and finds the one that reduces the residual sum of squares the most.
-   This is very similar to OLS, which finds coefficient values that minimises the sum of squared errors (also $\sum\limits_{i=1}^{n} (y_i - \hat{y}_i)$ )

Once we have split $x$, find the average $y$ for the observations in each region, and that is our prediction for the points in the region.

<br />

When there are multiple independent variables $x_1, x_2, ... , x_p$, regression trees will do the following:

1.  Test all $x$ variables and the possible split thresholds $s$ for each.
2.  Identify the specific $x$ variable $x_j$ and specific split threshold $s$ that results in the greatest reduction in the residual sum of squares.
3.  Now, divide chosen $x_j$ at chosen threshold $s$ to create two regions: $x_j < s, x_j ≥s$
4.  Now, we repeat the process of testing all $x$ variables (including the one we just chose) and the possible thresholds $s$ for each. HOWEVER, this time, instead of looking for this through the entire data set, look only inside one of the 2 regions we just created.
5.  Now, when we split again after finding the optimal $x$ and $j$, we only divide one of the 2 subregions we created in step 3.
6.  Continue dividing each subregions further, until we meet a stopping criterion (this is often that no region has more than 5 observations).
7.  For each final subregion, find the average $y$ value of the observations in that subregion, and that is our prediction for all points in that subregion.

<br />

Below is an example of a tree regression, with $y$ being **gini index**, $x_1$ is **export volume**, and $x_2$ is **tax percent**. Note the following:

-   The top division is a split in **taxpercent**. This tells use that this $x$ variable with the split point $x = 34.2935$ is the split that reduces the residual sum of squares the most. This is the most "influential" variable on the prediction.

-   Notice how after the top division, we go to divide the subregions. We can only further divide subregions already created (no going back to the whole data set).

-   Notice how the variable **taxpercent** appears again on the right hand side. This is possible for a variable to occur multiple times.

-   At the end, you see the numbers at the bottom. These are called **leaves**, and are the final categories of the tree with their mean of $y$ labelled.

```{r, echo = FALSE}

tree_example <- tree(gini ~ export + taxpercent, data = df)

plot(tree_example)
text(tree_example)
```

<br />

While regression trees are a cool way to visualise predictions, they tend to have high variance: just slightly changing the data will result in a completely different tree being generated.

-   This is an issue, especially if we are interested in predicting out-of-sample observations.

Thus, simple regression trees are almost never used in data science for predictions. The more advanced models of [Bagging and Random Forest](#bagging) that build on regression trees are frequently used and solve many of the issues with regression trees.

<br />

------------------------------------------------------------------------

[Table of Contents](#contents) \| [Course Homepage](https://politicalscience.github.io/ml)

# Regression Trees in R {#r}

To conduct Tree Regressions, wee need the package **trees**:

```{r, message = FALSE, eval = FALSE}
library(tree)
```

Remember to load tidyverse.

```{r, message = FALSE, eval = FALSE}
library(tidyverse)
```

Let us also load the dataset we will be using for these examples (feel free to load your own dataset)

```{r, message = FALSE, eval = FALSE}
df <- read_csv("voctaxdata.csv")
```

<br />

#### Creating the Model in R

Tree regressions in R are conducted using the **tree()** function. We can view the summary by simply printing the regression variable. We can view a tree plot with the **plot()** and **text()** functions. The syntax is as follows:

```{r, eval = FALSE}
# remember to load package tree

tree_reg<- tree(Y ~ X1 + X2, data = df)

# see written summary of tree
tree_reg

# create tree plot
plot(tree_example)
text(tree_example)
```

These are the parts of the syntax that can be altered:

-   **tree_reg** is the variable I am saving my regression model to. *You can name this anything you want to.*

-   **Y** is the Y variable (Dependent variable) you are trying to predict, and **X1, X2,...** are the X variables (independent variables) you are using to get your prediction. *Replace these with the variables you want to use.*

    -   NOTE: Always put the Y variable before the X variable. Separate the two with a tilda **\~**
    -   NOTE: You can add more simply by using a **+** sign and adding another variable.

-   **df** is the name of the data frame that I am drawing these X and Y variables from. *Replace this with the name of your data frame.*

<br />

#### Prediction in R

If we are interested in prediction, we can use the **predict()** function. You can predict in-sample data by setting **newdata** = the data frame you used for regression. You can predict out-of-sample data by using a dataframe with the same variables but new values. The syntax is as follows:

```{r, eval = FALSE}
#create new df for comparison of actual and prediction
df_results <- df %>%
  select(Y) #optional, may help with readability

# newdata is what values of X1, X2... to predict for.
df_results$prediction <- predict(tree_reg, newdata = df)

# brief glimpse of the results
head(df_results)
```

These are the parts of the syntax that can be altered:

-   **df_results** is the results data frame I am creating. *You can name this anything you want to.*

-   **Y** is the Y variable I am trying to predict. *Replace this with the name of your Y variable.*

-   **tree_reg** is the variable I am saved my prior model to. *Rename this to what your prior model was named.*

-   **df** is the name of the data frame that houses the $x$ values I want to predict for. *Replace this with the name of your data frame with the* $x$ *values you want to predict for.*

<br />

#### Example in R

Take this following example of a tree regression, with $y$ being **gini index**, $x_1$ is **export volume**, and $x_2$ is **tax percent** (same as the figure displayed earlier)

```{r}

# building a tree
tree_reg <- tree(gini ~ export + taxpercent, data = df)

# making predictions
# create new df for comparison of actual and prediction
df_results <- df %>%
  select(gini)

# predict in-sample data
df_results$prediction <- predict(tree_reg, newdata = df)

# brief glimpse of the results
head(df_results)
```

Here, you can see the predictions vs. the actual values in my dataframe **df_results**.

<br />

------------------------------------------------------------------------

[Table of Contents](#contents) \| [Course Homepage](https://politicalscience.github.io/ml)
