---
title: "Political Science Handbook"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Statistics/Econometrics with R for Political Science

This page is a handbook for statistics/econometrics in Political Science, using the R programming language, intended to help implement statistical/econometric methods in the R language. In fact, this website was written with R-Markdown!

What this handbook is NOT for:

-   This handbook is NOT a course on statistics or quantitative methods. This handbook will not go into the mathematics, statistics, and probability theory behind these methods. This handbook is designed for researchers who have a baseline understanding of statistics/econometrics, and want to apply those techniques in R. [Resources for Learning Statistics and R-Programming].

-   This handbook is NOT a course on the R language. It assumes you know basics, such as how libraries work, the pipe operator, objects, and general syntax. [Resources for Learning Statistics and R-Programming].

This handbook heavily relies on the tidyverse package. Things will not work without it. Install it in the following way:

```{r, eval = FALSE}
install.packages("tidyverse")
```

There are other packages that will be used, but are only applicable to specific tools, so I will introduce them when necessary.

[About the Creator]

<br />

------------------------------------------------------------------------

------------------------------------------------------------------------

# Chapter 0: Data Skills in R

*This section will teach you about handling data in R. This section isn't neccesary for understanding statistical applications in political science, but is useful as an introduction to R and data management.*

+------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------+
| Lesson                                                                             | Description                                                                                     |
+====================================================================================+=================================================================================================+
| [**Managing Data**](https://politicalscience.github.io/r/managing_data.html)       | Import data into R, selecting and filtering, adding columns, renaming columns, merging datasets |
+------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------+
| [**Visualizing Data**](https://politicalscience.github.io/r/visualizing_data.html) | Ggplot histograms, boxplots, and scatterplots with best fit, paris() function, correlations     |
+------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------+
| **Long and Wide Data**                                                             | Long vs. wide data, pivot longer, pivot wider                                                   |
+------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------+

<br />

------------------------------------------------------------------------

# Chapter 1: Regression Analysis

*This section will introduce you to the common regression methods used in the social sciences, which are the primary tools used for political science research.*

+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Lesson                                                                                             | Description                                                                                                                                                                     |
+====================================================================================================+=================================================================================================================================================================================+
| [**Single and Multivariate OLS Regression**](https://politicalscience.github.io/r/regression.html) | This lesson introduces the linear model of regression, which is used to predict continuous Y variables from independent X variables, and establish correlation between X and Y. |
|                                                                                                    |                                                                                                                                                                                 |
|                                                                                                    | [Topics]{.underline}: Simple OLS Regression, Graphing Simple Linear Regression, Categorical X Variables, Multivariate Linear Regression                                         |
+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Key Assumptions of OLS Regression**                                                              | How to deal with outliers, heteroskedasticity, multicollinearity, types of errors, residual plots                                                                               |
+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Moderating/Interaction Effect OLS Regression**                                                   | Categorical interaction/moderating effect, continuous interaction/moderating effect                                                                                             |
+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| [**Logistic Regression**](https://politicalscience.github.io/r/logistic.html)                      | Binomial logistic regression, multinomial logistic regression, ordinal logistic regression                                                                                      |
+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Poisson and Negative Binomial Regression**                                                       | Poisson regression, Negative Binomial Regression                                                                                                                                |
+----------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

<br />

------------------------------------------------------------------------

# Chapter 2: Causal Inference

*(Coming Winter 2024) Social sciences, including both Political Science and Economics, are heavily oriented around finding causal relationships between variables. This section will introduce how causality can be established in research design, with a primary focus on observational data and causality.*

+----------------------------------+---------------------------------------+
| Lesson                           | Description                           |
+==================================+=======================================+
| **Selection on Observables**     |                                       |
+----------------------------------+---------------------------------------+
| **Randomised Controlled Trials** |                                       |
+----------------------------------+---------------------------------------+
| **Instrumental Variables**       |                                       |
+----------------------------------+---------------------------------------+
| **Regression Discontinuity**     |                                       |
+----------------------------------+---------------------------------------+
| **Differences in Differences**   |                                       |
+----------------------------------+---------------------------------------+
| **Survey Experiments**           |                                       |
+----------------------------------+---------------------------------------+

<br />

------------------------------------------------------------------------

# Section 3: Supervised Learning Methods

*This section will introduce supervised learning prediction techniques from data science and machine learning. These more advanced methods, while less easy to interpret and draw causal relationships, are more accurate in making predictions.*

+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+
| Lesson                                                                               | Description                                                                                                             |
+======================================================================================+=========================================================================================================================+
| **Non-Linear Methods and GAMs**                                                      | Polynomial regressions, step functions, splines, local linear regression, generalised additive models.                  |
+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+
| [**Tree-Based Prediction Methods**](https://politicalscience.github.io/r/trees.html) | Bagging, random forest, importance plots.                                                                               |
+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+
| **Evaluating Prediction Methods**                                                    | Training/testing sets, k-fold cross validation, bootstrap sampling                                                      |
+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+
| **Classification Prediction**                                                        | Naive bayes, tree-Based classification, evaluation of classification models (confusion matrix, sensitivity/specificity) |
+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+
| **Regularization**                                                                   | Ridge regression, lasso regression, tuning grid for lambda parameter                                                    |
+--------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------+

<br />

------------------------------------------------------------------------

# Section 4: Unsupervised Learning Methods

*This section will introduce you to some unsupervised learning methods. These methods allow us to learn more about our data, even if we aren't sure what we want to predict.*

+---------------------------+---------------------------------------------+
| Col1                      | Col2                                        |
+===========================+=============================================+
| **Cluster Analysis**      | K-means clustering, hierarchical clustering |
+---------------------------+---------------------------------------------+
| **Dimensional Reduction** | Principle component analysis                |
+---------------------------+---------------------------------------------+

<br />

------------------------------------------------------------------------

# Section 5: Quantitative Textual Analysis

*This section will introduce you to new machine learning advances in turning textual data into quantitative data. These methods have become increasing popular and widespread in academic political science.*

+-------------------------------------+----------------------------------------------------------+
| Lesson                              | Description                                              |
+=====================================+==========================================================+
| F**eature Selection**               | Document feature matrix, bag of words, N-grams, trimming |
+-------------------------------------+----------------------------------------------------------+
| **Dictionary Methods**              |                                                          |
+-------------------------------------+----------------------------------------------------------+
| **Similarity Metrics**              |                                                          |
+-------------------------------------+----------------------------------------------------------+
| **Classification Methods for Text** | Naive Bayes for text,                                    |
+-------------------------------------+----------------------------------------------------------+
| **Topic Modelling**                 | Latent Dirichlet Allocation, Structural Topic Model      |
+-------------------------------------+----------------------------------------------------------+

<br />

------------------------------------------------------------------------

# Further Information

#### About the Creator

This handbook was created by Kevin Lingfeng Li, in R-Markdown!

-   [LinkedIn](https://www.linkedin.com/in/kevinlingfengli/) \| [Personal Website](https://kevinli03.github.io)

I completed my undergraduate in BSc Politics at King's College London, focusing on statistical methods and game theory, obtaining a First-Class Honours (US equivalent: Summa Cum Laude) classification. I then went on to pursue a MSc in Political Science & Political Economy at the London School of Economics (LSE).

I created this handbook as a helpful guide while studying for my Masters degree. I realised that I had learned a ton of methods, but could not remember the syntax for all of them. Online resources often over-complicated the functions needed for political science analysis. Thus, in my free time, I created this handbook as a quick guide to implementing common statistical techniques.

I do my best to update this handbook: it is a tool not only for my classmates and other political science students, but for me myself. However, as a full-time student with an internship, I may not be able to update this site at a fast enough pace.

<br />

------------------------------------------------------------------------

#### Resources for Learning Statistics and R-Programming

-   For those who need an introduction to statistics, I recommend the **Khan Academy** course on AP Stats for a mathematical introduction: <https://www.khanacademy.org/math/ap-statistics>

-   There is also a terrific free textbook on **Statistical learning with applications to R**: <https://www.statlearning.com>

-   **BurkeyAcademy** provides free online econometrics courses and implementations in R: <https://www.burkeyacademy.com/statistics-econometrics>

-   For more advanced individuals who want to learn more in-depth about statistics, see this excellent **MIT lecture series**: <https://www.youtube.com/playlist?list=PLUl4u3cNGP61ATaGTFcSp7bhogloD2wHP>

-   A free resource to learn R with is **CodeAcademy**: <https://www.codecademy.com/enrolled/courses/learn-r>

-   For those who have spare cash to spend, there is an excellent R programming course **(Data Analysis with R) offered by IBM on Coursera**, which teaches R and statistics: <https://www.coursera.org/learn/data-analysis-with-r>

-   I also highly recommend those who can afford it, to attend the **London School of Economic (LSE) Summer School**, which has an excellent course - **ME314 Introduction to Data Science and Machine Learning**. Many of these methods are covered in depth in this course, and the course is run by one of the developers of Quanteda, the main textual-analysis R package.

------------------------------------------------------------------------

The contents of this handbook are advisory: they should not be taken as gospel. The contents of this handbook are accurate to the best of my ability, as of August 2024. Later updates to R and packages may result in errors.
