---
title: "Political Science Handbook"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Handbook - Quantitative Methods in R

This page is a handbook for quantitative methods in Political Science, using the R programming language, intended to help implement statistical/econometric methods in the free R language. In fact, this website was written with R-Markdown!

What this handbook is NOT for:

-   This handbook is NOT a course on statistics or quantitative methods. This handbook will not go into the mathematics, statistics, and probability theory behind these methods. This handbook is designed for researchers who have an understanding of statistics/econometrics, and want to apply those techniques in R. [Resources for Learning Statistics and R-Programming].

-   This handbook is NOT a course on the R language. It assumes you know basics, such as how libraries work, the pipe operator, objects, and general syntax. [Resources for Learning Statistics and R-Programming].

[About the Creator]

------------------------------------------------------------------------

## Section 0: Key Info

This handbook heavily relies on the tidyverse package. Things will not work without it. Install it in the following way:

```{r, eval = FALSE}
install.packages("tidyverse")
```

There are other packages that will be used, but are only applicable to specific tools, so I will introduce them when necessary.

------------------------------------------------------------------------

## Section 1: Data

[Managing Data](https://politicalscience.github.io/r/managing_data.html)

-   Import data, select, filter, arrange (sort), add columns, rename columns, merging datasets

Visualising Data

-   Ggplot histograms, boxplots, and scatterplots with best fit, pairs() function & correlations

Long and Wide Data

-   Pivot longer, pivot wider

------------------------------------------------------------------------

## Section 2: Methods Toolkit

Linear Regression & Interaction

-   Simple & Multivariate Linear Regression, Interaction/Moderating Effect

Non-Linear Methods

-   Polynomial Regressions, Step Functions, Splines, Local Linear Regression, Generalized Additive Models

Tree-Based Methods

-   Bagging, Random Forest

Evaluating Prediction Methods

-   Training/testing sets, K-fold Cross Validation, Bootstrap sampling

Classification Methods

-   Logistic Regression, Naive Bayes, Evaluation of Classification Models (Confusion Matrix, Sensitivity/Specificity)

Regularization

-   Ridge, Lasso, Tuning grid for Lambda

Dimensional Reduction

-   Principle Component Analysis

Cluster Analysis

-   K-Means Clustering, Hierarchical Clustering

------------------------------------------------------------------------

## Section 3: Textual Methods Toolkit

Feature Selection

-   Document Feature Matrix, Bag of Words, N-Grams, Trimming

Dictionary Methods

Similarity Metrics

Classification Methods for Text

-   Naive Bayes Classification for Text

Topic Modelling

-   Latent Dirichlet Allocation, Structural Topic Model

------------------------------------------------------------------------

# Further Information

#### About the Creator

This handbook was created by Kevin Lingfeng Li, in R-Markdown!

I completed his undergraduate in BSc Politics at King's College London, focusing on statistical methods and game theory, obtaining a First-Class Honours classification. I then went on to pursue a MSc in Political Science & Political Economy at the London School of Economics (LSE).

I created this handbook as a helpful guide while studying for his Masters degree. I realised that he had learned a ton of methods, but could not remember the syntax for all of them. Online resources often overcomplicated the functions needed for political science analysis. Thus, in my free time, I created this handbook as a quick guide to implementing common statistical techniques.

I do my best to update this handbook: it is a tool not only for my classmates and other political science students, but for me myself. However, as a full-time student with an internship, I may not be able to update this site at a fast enough pace.

------------------------------------------------------------------------

#### Resources for Learning Statistics and R-Programming

-   For those who need an introduction to statistics, I recommend the **Khan Academy** course on AP Stats for a mathematical introduction: <https://www.khanacademy.org/math/ap-statistics>

-   There is also a terrific free textbook on **Statistical learning with applications to R**: <https://www.statlearning.com>

-   For more advanced individuals who want to learn more in-depth about statistics, see this excellent **MIT lecture series**: <https://www.youtube.com/playlist?list=PLUl4u3cNGP61ATaGTFcSp7bhogloD2wHP>

-   A free resource to learn R with is **CodeAcademy**: <https://www.codecademy.com/enrolled/courses/learn-r>

-   For those who have spare cash to spend, there is an excellent R programming course **(Data Analysis with R) offered by IBM on Coursera**, which teaches R and statistics: <https://www.coursera.org/learn/data-analysis-with-r>

-   I also highly recommend those who can afford it, to attend the **London School of Economic (LSE) Summer Schoo**l, which has an excellent course - **ME314 Introduction to Data Science and Machine Learning**. Many of these methods are covered in depth in this course, and the course is run by one of the developed of Quanteda, the main textual-analysis R package.

------------------------------------------------------------------------

The contents of this handbook are advisory: they should not be taken as gospel. The contents of this handbook are accurate to the best of my ability, as of August 2024. Later updates to R and packages may result in errors.
